---
title: "Medium Task for GSOC"
author: "Javad Vahdat Atashgah"
format: html
toc: true
toc-depth: 4
toc-location: left
engine: knitr
execute: 
  warning: false
  message: false
---

## Task 2 (Medium)

- Use the `GSODR` package to download one year of daily weather data (temperature and precipitation) for a station in Victoria, Australia, near where large numbers of platypus are spotted.

### Actionable Workflow

My approach is doing a spatial join between platypus sightings and GSOD weather stations, then aggregating to identify the most relevant stations.

#### Filter Platypus Sightings to Victoria

Load Data platypus
```{r}
load(file= "../data/platypus_2024.RDA")
```

Filter Platypus Sightings to Victoria
```{r}
library(dplyr)
platypus_2024_vic <- platypus_2024 %>%
  filter(decimalLatitude >= -39, decimalLatitude <= -34,
         decimalLongitude >= 141, decimalLongitude <= 150)
```

### Get All GSOD Stations in Victoria

Install and load package
```{r}
if (!requireNamespace("GSODR", quietly = TRUE)) {install.packages("GSODR")}

library(GSODR)
```

Download and Return weather station Data
```{r}
stations <- get_inventory()
```

```{r}
vic_stations <- stations %>%
  filter(CTRY == "AS", LAT >= -39, LAT <= -33.9, LON >= 140.9, LON <= 150.1) %>%
  select(STNID, NAME, LAT, LON)

head(vic_stations)
```

### Relate Each Occurrence to Its Nearest Station
Use a spatial distance function to map each platypus record to its closest station:

```{r}
if (!requireNamespace("geosphere", quietly = TRUE)) {install.packages("geosphere")}

library(geosphere)

# For each occurrence, find the nearest station
nearest_station_ids <- sapply(1:nrow(platypus_2024_vic), function(i) {
  dists <- distHaversine(
    cbind(vic_stations$LON, vic_stations$LAT),
    c(platypus_2024_vic$decimalLongitude[i],    platypus_2024_vic$decimalLatitude[i])
  )
  vic_stations$STNID[which.min(dists)]
})

# Add the nearest station ID to the platypus data
platypus_2024_vic$nearest_station <- nearest_station_ids

```


### Count Number of Sightings per Station

```{r}
station_counts <- platypus_2024_vic %>%
  count(nearest_station, sort = TRUE) %>%
  left_join(vic_stations, by = c("nearest_station" = "STNID")) |> 
  distinct()

head(station_counts)
```

```{r}
#install.packages("ozmaps")

library(ggplot2)
library(ozmaps)


# Filter only VIC map
vic_map <- ozmaps::ozmap_states %>% 
  dplyr::filter(NAME == "Victoria")

# Plot
ggplot() +
  geom_sf(data = vic_map, fill = "gray95", color = "gray70") +
  geom_point(data = station_counts, 
             aes(x = LON, y = LAT, size = n, color = n), 
             alpha = 0.8) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Platypus Observations per Weather Station in Victoria",
    x = "Longitude", y = "Latitude",
    size = "Count", color = "Count"
  ) +
  coord_sf() +
  theme_minimal()
```

### Choose Top Stations

```{r}
top_stations <- station_counts %>%
  slice_max(n, n = 3)  # Top 3 stations by sighting proximity
head(top_stations)
```

### Download Weather Data for Top 3 Stations

```{r}
weather_data_list <- lapply(top_stations$nearest_station, function(stn) {
  get_GSOD(years = 2024, station = stn)
})

# Combine if needed
weather_data_combined <- dplyr::bind_rows(weather_data_list)

```

### tidying datasets colnames
```{r, echo=FALSE}
platypus_2024_vic |> transmute(
  LAT = decimalLatitude,
  LON = decimalLongitude,
  YEAR = year,
  eventDate = eventDate,
  scientificName = scientificName,
  STNID = nearest_station
) -> platypus_2024_vic
  
weather_data_combined |> transmute(
  STNID = STNID,
  NAME = NAME,
  LAT = LATITUDE,
  LON = LONGITUDE,
  DATE = YEARMODA,
  YEAR = YEAR,
  MONTH = MONTH,
  DAY = DAY,
  YDAY = YDAY,
  TEMP = TEMP,
  TEMP_MIN = MIN,
  TEMP_MAX = MAX,
  PRCP = PRCP,
  DEWP = DEWP,
  WDSP = WDSP,
  VISIB = VISIB
) -> weather_data_combined
```



### save data

```{r}
save(platypus_2024_vic, file = "../data/platypus_2024_vic.RDA")

save(vic_stations, file = "../data/vic_stations.RDA")

save(weather_data_combined, file = "../data/weather_data.RDA")
```


### 🔗 Data Structure and Relationships

All three datasets share a common key: **`STNID`**, which represents a unique weather station ID.

#### 1. `platypus_2024_vic`

| Column           | Description                                 |
|------------------|---------------------------------------------|
| `LAT`, `LON`     | Coordinates of platypus sightings           |
| `YEAR`, `eventDate` | Year and full timestamp of sighting      |
| `scientificName` | Name of the species (platypus)             |
| `STNID`          | Nearest weather station ID (linked key)     |

#### 2. `vic_stations`

| Column    | Description                        |
|-----------|------------------------------------|
| `STNID`   | Weather station ID (primary key)   |
| `NAME`    | Name of the station                |
| `LAT`, `LON` | Coordinates of the station     |

#### 3. `weather_data_combined`

| Column    | Description                                 |
|-----------|---------------------------------------------|
| `STNID`   | Weather station ID (linked key)             |
| `DATE`    | Date of observation                         |
| `TEMP`    | Daily mean temperature (°C)                 |
| `TEMP_MIN`, `TEMP_MAX` | Daily min/max temperatures     |
| `PRCP`    | Daily precipitation (mm)                    |
| `DEWP`, `WDSP`, `VISIB` | Other daily weather variables |

---

These datasets can be **joined using `STNID`** for analyses like:

- Matching platypus sightings with same-day weather
- Analyzing seasonal patterns in specific regions
- Aggregating sightings and climate trends by station
